disaggregation: An \proglang{R} package for Bayesian spatial disaggregation modelling

### 2019-12-03 article.Rnw
Edits to the model section

For the disaggregation model, we first model the rate at pixel level, with the rate in pixel $j$ of polygon $i$ given by:
%
\begin{equation} \label{eq:lp}
\textrm{link}(\textrm{rate}_{ij}) = \beta_0 + \beta X_{ij} + \textrm{GP}(s_{ij}) + u_i 
\end{equation}
%
Here $\beta$ are regression coefficients, $X_{ij}$ are covariate values, GP is a Gaussian random field and $u_i$ is a polygon-specific effect. The user defined link function is typically log, identity and logit for Poisson, Normal and Binomial likelihoods, respectively  The Gaussian random field has a Matern covariance function parameterised by $\rho$, the range (approximately the distance beyond which correlation is less than 0.1), and $\sigma$, the marginal standard deviation. As we are working in a Bayesian setting, each of the model parameters and hyperparameters are given a prior, which is discussed later.

@@

%
\begin{itemize}
\item {\textbf{Gaussian}
\begin{equation} \label{eq:gaussian}
y_i \sim \textrm{Normal}(\textrm{cases}_i, \sigma_i)
\end{equation}
Here $\sigma_i = \sigma \sqrt{\sum_j a_{ij}^2}$, where  $\sigma$ is the pixel-level dispersion (a parameter learnt by the model) }

\item \textbf{Binomial}
\begin{equation} \label{eq:binomial}
y_i \sim \textrm{Binomial}(M_i, \textrm{rate}_i)
\end{equation}
For a survey in polygon $i$, $y_i$ is the number positive cases and $M_i$ is the number tested.

\item \textbf{Poisson}
\begin{equation} \label{eq:poisson}
y_i \sim \textrm{Pois}(\textrm{cases}_i)
\end{equation}

\end{itemize}
%

Note the normal likelihood is modelled in rate space as it is rarely reasonable to assume that the dispersion of the normal likelihood is equal for polygons containing different numbers of cases. Also, the binomial likelihood can be used for survey data, for example when the response data consists of the number of people who tested positive for a disease and the total number of people that were tested.

>>>>

For the disaggregation model we model the rate at pixel level, with the likelihood for then observed data then given by aggregating these pixel level rates. The rate in pixel $j$ of polygon $i$ at location $s_{ij}$ is given by:
%
\begin{equation} \label{eq:lp}
\textrm{link}(\textrm{rate}_{ij}) = \beta_0 + \beta X_{ij} + \textrm{GP}(s_{ij}) + u_i 
\end{equation}
%
where $\beta$ are regression coefficients, $X_{ij}$ are covariate values, GP is a Gaussian random field and $u_i$ is a polygon-specific effect. The user defined link function is typically log, identity and logit for Poisson, Normal and Binomial likelihoods, respectively  The Gaussian random field has a Matern covariance function parameterised by $\rho$, the range (approximately the distance beyond which correlation is less than 0.1), and $\sigma$, the marginal standard deviation. As we are working in a Bayesian setting, each of the model parameters and hyperparameters are given a prior, which is discussed later.

@@

%
\begin{itemize}
\item \textbf{Poisson}
\begin{equation} \label{eq:poisson}
y_i \sim \textrm{Pois}(\textrm{cases}_i)
\end{equation}

\item {\textbf{Gaussian}
\begin{equation} \label{eq:gaussian}
y_i \sim \textrm{Normal}(\textrm{cases}_i, \sigma_i)
\end{equation}
Here $\sigma_i = \sigma \sqrt{\sum_j a_{ij}^2}$, where  $\sigma$ is the pixel-level dispersion (a parameter learnt by the model) }

\item \textbf{Binomial}
\begin{equation} \label{eq:binomial}
y_i \sim \textrm{Binomial}(M_i, \textrm{rate}_i)
\end{equation}

\end{itemize}
%
In the example of disease mapping, Poisson or Gaussian likelihoods could be used when the quantity observed, $y_i$, is the total number of cases in a given polygon. The Binomial model could be used when $y_i$ is the prevalence of a disease in a sample of $M_i$ people in the polygon.
